---
title: "Predicting Ship Fuel Efficiency Class: A Machine Learning Classification Approach"
author: "Shipping Performance Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6,
  fig.align  = "center"
)
```

---

# Libraries

```{r libraries}
library(tidyverse)
library(ggplot2)
library(caret)
library(randomForest)
library(ranger)
library(gbm)
library(nnet)
library(kknn)
library(pROC)
library(MLmetrics)
library(knitr)
library(kableExtra)
library(scales)
library(viridis)
library(gridExtra)
library(RColorBrewer)
```

---

# 1. Problem Statement

**Objective:** Predict a voyage's **fuel efficiency class** (Low / Medium / High) from
its operational and vessel characteristics — a 3-class classification problem.

**Why this target?**
Diagnostic testing confirmed that `Maintenance_Status` was randomly assigned in this
synthetic dataset (ANOVA F < 2.0 across all features). `Efficiency_nm_per_kWh` is a
real measured outcome that operational features such as speed, engine power, cargo load,
and route type can logically influence.

**Target variable:** `Efficiency_Class` — derived by binning `Efficiency_nm_per_kWh`
into equal thirds:

| Class | Efficiency Range | Meaning |
|-------|-----------------|---------|
| `Low`    | Bottom 33% | Poor fuel economy — high operating cost per nm |
| `Medium` | Middle 33%  | Average fuel economy |
| `High`   | Top 33%     | Excellent fuel economy — low cost per nm |

> `Efficiency_nm_per_kWh` is **excluded from the feature set** — it is the source
> of the target and would cause data leakage if included.

**Business value:** Predicting efficiency class before departure allows operators to
adjust speed, load, and routing to target the High class and reduce fuel costs.

---

# 2. Data Loading & Preprocessing

## 2.1 Load Data

```{r load}
df <- read.csv("Ship_Performance_Dataset.csv", stringsAsFactors = FALSE)
df$Date <- as.Date(df$Date)

cat("Dimensions:", nrow(df), "rows ×", ncol(df), "cols\n")
```

## 2.2 Create Target — Efficiency_Class

```{r create-target}
df <- df %>%
  mutate(
    Efficiency_Class = factor(
      ntile(Efficiency_nm_per_kWh, 3),
      levels = c(1, 2, 3),
      labels = c("Low", "Medium", "High")
    )
  )

cat("Efficiency_Class distribution:\n")
print(table(df$Efficiency_Class))
cat("\nProportions (%):\n")
print(round(prop.table(table(df$Efficiency_Class)) * 100, 1))
```

```{r eff-dist-plot, fig.height=4}
# Efficiency distribution coloured by class
ggplot(df, aes(x = Efficiency_nm_per_kWh, fill = Efficiency_Class)) +
  geom_histogram(bins = 50, color = "white", alpha = 0.9) +
  scale_fill_manual(values = c("Low" = "#e74c3c", "Medium" = "#f39c12", "High" = "#2ecc71")) +
  labs(title = "Efficiency_nm_per_kWh Distribution — Coloured by Derived Class",
       subtitle = "Three equal-frequency bins (ntile = 3)",
       x = "Efficiency (nm/kWh)", y = "Count", fill = "Efficiency Class") +
  theme_minimal(base_size = 13)
```

```{r class-dist-plot, fig.height=4}
class_df <- as.data.frame(table(df$Efficiency_Class)) %>%
  rename(Class = Var1, Count = Freq) %>%
  mutate(Pct = round(Count / sum(Count) * 100, 1))

ggplot(class_df, aes(x = Class, y = Count, fill = Class)) +
  geom_col(color = "white", width = 0.65) +
  geom_text(aes(label = paste0(Count, "\n(", Pct, "%)")),
            vjust = -0.3, fontface = "bold", size = 4) +
  scale_fill_manual(values = c("Low" = "#e74c3c", "Medium" = "#f39c12", "High" = "#2ecc71")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Target Class Distribution — Efficiency Class",
       subtitle = "Perfectly balanced by construction (ntile) — No SMOTE required",
       x = "Efficiency Class", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

## 2.3 Feature-Target Relationship Check

Before modelling, verify that features have real signal for this target using ANOVA F-statistics.

```{r signal-check, fig.height=5}
num_cols <- c("Speed_Over_Ground_knots", "Engine_Power_kW", "Distance_Traveled_nm",
              "Draft_meters", "Cargo_Weight_tons", "Operational_Cost_USD",
              "Revenue_per_Voyage_USD", "Turnaround_Time_hours",
              "Seasonal_Impact_Score", "Weekly_Voyage_Count", "Average_Load_Percentage")

fstats <- sapply(num_cols, function(col) {
  m <- aov(as.formula(paste(col, "~ Efficiency_Class")), data = df)
  summary(m)[[1]][["F value"]][1]
})

fstat_df <- data.frame(
  Feature   = num_cols,
  F_Stat    = round(fstats, 2),
  Signal    = ifelse(fstats > 10, "Strong", ifelse(fstats > 3, "Moderate", "Weak"))
)

ggplot(fstat_df, aes(x = reorder(Feature, F_Stat), y = F_Stat, fill = Signal)) +
  geom_col(color = "white", width = 0.7) +
  geom_hline(yintercept = 3,  linetype = "dashed", color = "#e74c3c",  linewidth = 0.8) +
  geom_hline(yintercept = 10, linetype = "dashed", color = "#27ae60", linewidth = 0.8) +
  annotate("text", x = 1.5, y = 4,  label = "Moderate threshold (F=3)",  color = "#e74c3c",  size = 3.5) +
  annotate("text", x = 1.5, y = 11, label = "Strong threshold (F=10)",   color = "#27ae60", size = 3.5) +
  scale_fill_manual(values = c("Strong" = "#2ecc71", "Moderate" = "#f39c12", "Weak" = "#e74c3c")) +
  coord_flip() +
  labs(title = "ANOVA F-Statistics — Feature Signal Strength for Efficiency_Class",
       subtitle = "Higher F = stronger relationship between feature and target",
       x = NULL, y = "F-Statistic", fill = "Signal") +
  theme_minimal(base_size = 12)

kable(fstat_df[order(-fstat_df$F_Stat), ],
      caption = "Feature Signal Strength (ANOVA F-Stat vs Efficiency_Class)",
      row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  column_spec(3, bold = TRUE,
              color = ifelse(fstat_df[order(-fstat_df$F_Stat),]$Signal == "Strong", "darkgreen",
                      ifelse(fstat_df[order(-fstat_df$F_Stat),]$Signal == "Moderate", "darkorange", "red")))
```

## 2.4 Build Feature Matrix

```{r preprocess}
# Drop Date, Efficiency_nm_per_kWh (source of target = data leakage)
df_model <- df %>%
  select(-Date, -Efficiency_nm_per_kWh, -Maintenance_Status) %>%
  mutate(
    Ship_Type         = factor(Ship_Type),
    Route_Type        = factor(Route_Type),
    Engine_Type       = factor(Engine_Type),
    Weather_Condition = factor(Weather_Condition)
  ) %>%
  drop_na()

cat("Model dataset:", nrow(df_model), "rows ×", ncol(df_model), "cols\n")
cat("Target distribution:\n")
print(table(df_model$Efficiency_Class))
```

```{r feature-list}
feature_cols <- setdiff(names(df_model), "Efficiency_Class")
cat("Features (", length(feature_cols), "):\n")
cat(paste(" •", feature_cols, collapse = "\n"), "\n")
```

---

# 3. Train / Test Split

**Strategy:** Stratified 80/20 split — preserves class proportions in both sets.

```{r split}
set.seed(42)
train_idx <- createDataPartition(df_model$Efficiency_Class, p = 0.80, list = FALSE)

train_df <- df_model[ train_idx, ]
test_df  <- df_model[-train_idx, ]

cat("Training set:", nrow(train_df), "rows\n")
cat("Test set:    ", nrow(test_df),  "rows\n\n")

cat("Train class balance:\n")
print(round(prop.table(table(train_df$Efficiency_Class)) * 100, 1))
cat("\nTest class balance:\n")
print(round(prop.table(table(test_df$Efficiency_Class)) * 100, 1))
```

---

# 4. Cross-Validation Setup

**Strategy:** 5-fold stratified cross-validation on the training set.
All models share the same CV folds for a fair comparison.

```{r cv-setup}
cv_ctrl <- trainControl(
  method          = "cv",
  number          = 5,
  classProbs      = TRUE,
  summaryFunction = multiClassSummary,
  savePredictions = "final",
  verboseIter     = FALSE
)

cat("CV: 5-fold stratified | Metric: Accuracy + Macro F1\n")
cat("Baseline (random chance): 33.3%\n")
```

---

# 5. Algorithm 1 — Multinomial Logistic Regression

**Baseline model.** Linear decision boundaries via softmax function.

```{r lr-train}
set.seed(42)
lr_model <- train(
  Efficiency_Class ~ .,
  data       = train_df,
  method     = "multinom",
  trControl  = cv_ctrl,
  preProcess = c("center", "scale"),
  trace      = FALSE
)

cat("Logistic Regression — Best CV Accuracy:",
    round(max(lr_model$results$Accuracy), 4), "\n")
```

```{r lr-predict}
lr_pred  <- predict(lr_model, test_df)
lr_probs <- predict(lr_model, test_df, type = "prob")

lr_cm <- confusionMatrix(lr_pred, test_df$Efficiency_Class)
cat("=== Logistic Regression — Test Set ===\n")
print(lr_cm$overall[c("Accuracy", "Kappa")])
cat("\nPer-class metrics:\n")
print(round(lr_cm$byClass[, c("Precision", "Recall", "F1")], 4))
```

---

# 6. Algorithm 2 — K-Nearest Neighbours (KNN)

**Instance-based learner.** Classifies by majority vote among K nearest neighbours.

```{r knn-train}
set.seed(42)
knn_model <- train(
  Efficiency_Class ~ .,
  data       = train_df,
  method     = "kknn",
  trControl  = cv_ctrl,
  preProcess = c("center", "scale"),
  tuneGrid   = expand.grid(
    kmax     = c(5, 7, 9, 11, 15),
    distance = 2,
    kernel   = "optimal"
  )
)

cat("KNN — Best K:", knn_model$bestTune$kmax, "\n")
cat("KNN — Best CV Accuracy:", round(max(knn_model$results$Accuracy), 4), "\n")
```

```{r knn-tune-plot, fig.height=4}
plot(knn_model, main = "KNN — Accuracy vs K (5-fold CV)")
```

```{r knn-predict}
knn_pred  <- predict(knn_model, test_df)
knn_probs <- predict(knn_model, test_df, type = "prob")

knn_cm <- confusionMatrix(knn_pred, test_df$Efficiency_Class)
cat("=== KNN — Test Set ===\n")
print(knn_cm$overall[c("Accuracy", "Kappa")])
cat("\nPer-class metrics:\n")
print(round(knn_cm$byClass[, c("Precision", "Recall", "F1")], 4))
```

---

# 7. Algorithm 3 — Random Forest

**Ensemble of decision trees** using bootstrap aggregation + random feature subsets.

```{r rf-train}
set.seed(42)
rf_model <- train(
  Efficiency_Class ~ .,
  data       = train_df,
  method     = "ranger",
  trControl  = cv_ctrl,
  preProcess = c("center", "scale"),
  importance = "impurity",
  tuneGrid   = expand.grid(
    mtry          = c(3, 5, 7, 9),
    splitrule     = "gini",
    min.node.size = c(1, 5)
  )
)

cat("Random Forest — Best mtry:", rf_model$bestTune$mtry, "\n")
cat("Random Forest — Best min.node.size:", rf_model$bestTune$min.node.size, "\n")
cat("Random Forest — Best CV Accuracy:", round(max(rf_model$results$Accuracy), 4), "\n")
```

```{r rf-predict}
rf_pred  <- predict(rf_model, test_df)
rf_probs <- predict(rf_model, test_df, type = "prob")

rf_cm <- confusionMatrix(rf_pred, test_df$Efficiency_Class)
cat("=== Random Forest — Test Set ===\n")
print(rf_cm$overall[c("Accuracy", "Kappa")])
cat("\nPer-class metrics:\n")
print(round(rf_cm$byClass[, c("Precision", "Recall", "F1")], 4))
```

```{r rf-importance, fig.height=6}
plot(varImp(rf_model), top = 15,
     main = "Random Forest — Top 15 Feature Importances")
```

---

# 8. Algorithm 4 — Gradient Boosting Machine (GBM)

**Sequential tree boosting.** Each tree corrects the residual errors of all prior trees
via gradient descent in function space.

```{r gbm-train}
set.seed(42)
gbm_model <- train(
  Efficiency_Class ~ .,
  data       = train_df,
  method     = "gbm",
  trControl  = cv_ctrl,
  preProcess = c("center", "scale"),
  verbose    = FALSE,
  tuneGrid   = expand.grid(
    n.trees           = c(100, 200),
    interaction.depth = c(3, 5),
    shrinkage         = c(0.05, 0.1),
    n.minobsinnode    = 10
  )
)

cat("GBM — Best n.trees:", gbm_model$bestTune$n.trees, "\n")
cat("GBM — Best interaction.depth:", gbm_model$bestTune$interaction.depth, "\n")
cat("GBM — Best shrinkage:", gbm_model$bestTune$shrinkage, "\n")
cat("GBM — Best CV Accuracy:", round(max(gbm_model$results$Accuracy), 4), "\n")
```

```{r gbm-predict}
gbm_pred  <- predict(gbm_model, test_df)
gbm_probs <- predict(gbm_model, test_df, type = "prob")

gbm_cm <- confusionMatrix(gbm_pred, test_df$Efficiency_Class)
cat("=== GBM — Test Set ===\n")
print(gbm_cm$overall[c("Accuracy", "Kappa")])
cat("\nPer-class metrics:\n")
print(round(gbm_cm$byClass[, c("Precision", "Recall", "F1")], 4))
```

```{r gbm-importance, fig.height=6}
plot(varImp(gbm_model), top = 15,
     main = "GBM — Top 15 Feature Importances")
```

---

# 9. Model Comparison

## 9.1 Macro F1-Score Table

```{r macro-f1}
macro_f1 <- function(cm) round(mean(cm$byClass[, "F1"], na.rm = TRUE), 4)

results_df <- data.frame(
  Algorithm     = c("Logistic Regression", "KNN", "Random Forest", "GBM"),
  CV_Accuracy   = round(c(max(lr_model$results$Accuracy),
                           max(knn_model$results$Accuracy),
                           max(rf_model$results$Accuracy),
                           max(gbm_model$results$Accuracy)), 4),
  Test_Accuracy = round(c(lr_cm$overall["Accuracy"],
                           knn_cm$overall["Accuracy"],
                           rf_cm$overall["Accuracy"],
                           gbm_cm$overall["Accuracy"]), 4),
  Kappa         = round(c(lr_cm$overall["Kappa"],
                           knn_cm$overall["Kappa"],
                           rf_cm$overall["Kappa"],
                           gbm_cm$overall["Kappa"]), 4),
  Macro_F1      = c(macro_f1(lr_cm), macro_f1(knn_cm),
                    macro_f1(rf_cm), macro_f1(gbm_cm))
)

best_row <- which.max(results_df$Macro_F1)

kable(results_df, caption = "Model Comparison — All Metrics (Baseline = 33.3%)") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = TRUE) %>%
  row_spec(best_row, bold = TRUE, background = "#d5f5e3") %>%
  column_spec(5, bold = TRUE)
```

## 9.2 Accuracy Comparison Plot

```{r comparison-plot, fig.height=5}
results_long <- results_df %>%
  pivot_longer(cols = c(CV_Accuracy, Test_Accuracy, Macro_F1),
               names_to = "Metric", values_to = "Score")

ggplot(results_long, aes(x = reorder(Algorithm, Score), y = Score, fill = Metric)) +
  geom_col(position = "dodge", color = "white", width = 0.7) +
  geom_hline(yintercept = 0.333, linetype = "dashed",
             color = "#e74c3c", linewidth = 0.8) +
  annotate("text", x = 0.7, y = 0.345, label = "Random baseline (33.3%)",
           color = "#e74c3c", size = 3.5) +
  geom_text(aes(label = round(Score, 3)),
            position = position_dodge(width = 0.7),
            hjust = -0.1, size = 3, fontface = "bold") +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(limits = c(0, 1.1), labels = percent_format()) +
  coord_flip() +
  labs(title = "Model Comparison — CV Accuracy vs Test Accuracy vs Macro F1",
       x = NULL, y = "Score", fill = "Metric") +
  theme_minimal(base_size = 13)
```

## 9.3 CV Resampling Distribution

```{r cv-resample-plot, fig.height=5}
resamps <- resamples(list(LR  = lr_model, KNN = knn_model,
                          RF  = rf_model, GBM = gbm_model))

bwplot(resamps, metric = "Accuracy",
       main = "5-Fold CV Accuracy Distribution — All Models")
```

---

# 10. Confusion Matrices

```{r confusion-matrices, fig.height=10}
plot_cm <- function(cm_obj, title, palette) {
  cm_tbl <- as.data.frame(cm_obj$table)
  names(cm_tbl) <- c("Predicted", "Actual", "Freq")
  ggplot(cm_tbl, aes(x = Actual, y = Predicted, fill = Freq)) +
    geom_tile(color = "white", linewidth = 0.8) +
    geom_text(aes(label = Freq), fontface = "bold", size = 5) +
    scale_fill_gradient(low = "white", high = palette) +
    labs(title = title, x = "Actual", y = "Predicted") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none")
}

p_lr  <- plot_cm(lr_cm,  "Logistic Regression", "#3498db")
p_knn <- plot_cm(knn_cm, "KNN",                 "#2ecc71")
p_rf  <- plot_cm(rf_cm,  "Random Forest",        "#9b59b6")
p_gbm <- plot_cm(gbm_cm, "GBM",                 "#e67e22")

grid.arrange(p_lr, p_knn, p_rf, p_gbm, ncol = 2,
             top = "Confusion Matrices — Test Set (Rows = Predicted, Cols = Actual)")
```

---

# 11. Per-Class Metrics

```{r per-class-metrics, fig.height=7}
extract_metrics <- function(cm, algo_name) {
  as.data.frame(cm$byClass[, c("Precision","Recall","F1")]) %>%
    rownames_to_column("Class") %>%
    mutate(Class     = str_remove(Class, "Class: "),
           Algorithm = algo_name)
}

per_class <- bind_rows(
  extract_metrics(lr_cm,  "Logistic Regression"),
  extract_metrics(knn_cm, "KNN"),
  extract_metrics(rf_cm,  "Random Forest"),
  extract_metrics(gbm_cm, "GBM")
) %>%
  pivot_longer(cols = c(Precision, Recall, F1),
               names_to = "Metric", values_to = "Value")

ggplot(per_class, aes(x = Algorithm, y = Value, fill = Algorithm)) +
  geom_col(position = "dodge", color = "white") +
  geom_text(aes(label = round(Value, 2)),
            vjust = -0.3, size = 2.8, fontface = "bold") +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(limits = c(0, 1.15), labels = percent_format()) +
  facet_grid(Class ~ Metric) +
  labs(title = "Per-Class Precision / Recall / F1 by Algorithm",
       x = NULL, y = "Score") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none")
```

---

# 12. ROC Curves (One-vs-Rest)

```{r roc-curves, fig.height=8}
classes <- levels(test_df$Efficiency_Class)

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

for (cls in classes) {
  actual_bin <- as.numeric(test_df$Efficiency_Class == cls)
  roc_lr  <- roc(actual_bin, lr_probs[[cls]],  quiet = TRUE)
  roc_knn <- roc(actual_bin, knn_probs[[cls]], quiet = TRUE)
  roc_rf  <- roc(actual_bin, rf_probs[[cls]],  quiet = TRUE)
  roc_gbm <- roc(actual_bin, gbm_probs[[cls]], quiet = TRUE)

  plot(roc_lr, col = "#3498db", lwd = 2,
       main = paste("ROC —", cls, "vs Rest"),
       xlab = "False Positive Rate", ylab = "True Positive Rate")
  lines(roc_knn, col = "#2ecc71", lwd = 2)
  lines(roc_rf,  col = "#9b59b6", lwd = 2)
  lines(roc_gbm, col = "#e67e22", lwd = 2)
  abline(0, 1, lty = 2, col = "grey60")
  legend("bottomright",
         legend = c(paste("LR  AUC =", round(auc(roc_lr),  3)),
                    paste("KNN AUC =", round(auc(roc_knn), 3)),
                    paste("RF  AUC =", round(auc(roc_rf),  3)),
                    paste("GBM AUC =", round(auc(roc_gbm), 3))),
         col = c("#3498db","#2ecc71","#9b59b6","#e67e22"),
         lwd = 2, cex = 0.75, bty = "n")
}

auc_df <- data.frame(
  Class = classes,
  LR  = sapply(classes, function(c) round(auc(roc(as.numeric(test_df$Efficiency_Class==c), lr_probs[[c]],  quiet=TRUE)), 4)),
  KNN = sapply(classes, function(c) round(auc(roc(as.numeric(test_df$Efficiency_Class==c), knn_probs[[c]], quiet=TRUE)), 4)),
  RF  = sapply(classes, function(c) round(auc(roc(as.numeric(test_df$Efficiency_Class==c), rf_probs[[c]],  quiet=TRUE)), 4)),
  GBM = sapply(classes, function(c) round(auc(roc(as.numeric(test_df$Efficiency_Class==c), gbm_probs[[c]], quiet=TRUE)), 4))
)
par(mfrow = c(1,1))
```

```{r auc-table}
kable(auc_df, caption = "AUC per Class (One-vs-Rest) — All Models") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  column_spec(which.max(colMeans(auc_df[,-1])) + 1, bold = TRUE, background = "#d5f5e3")
```

---

# 13. Feature Importance Comparison

```{r feature-importance-compare, fig.height=7}
rf_imp_df <- varImp(rf_model)$importance %>%
  rownames_to_column("Feature") %>%
  mutate(Importance = rowMeans(dplyr::select(., -Feature)),
         Algorithm  = "Random Forest") %>%
  dplyr::select(Feature, Importance, Algorithm) %>%
  slice_max(Importance, n = 10)

gbm_imp_df <- varImp(gbm_model)$importance %>%
  rownames_to_column("Feature") %>%
  mutate(Importance = rowMeans(dplyr::select(., -Feature)),
         Algorithm  = "GBM") %>%
  dplyr::select(Feature, Importance, Algorithm) %>%
  slice_max(Importance, n = 10)

imp_combined <- bind_rows(rf_imp_df, gbm_imp_df)

ggplot(imp_combined,
       aes(x = reorder(Feature, Importance), y = Importance, fill = Algorithm)) +
  geom_col(color = "white", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("Random Forest" = "#9b59b6", "GBM" = "#e67e22")) +
  facet_wrap(~ Algorithm, scales = "free_x") +
  labs(title = "Top 10 Feature Importances — Random Forest vs GBM",
       x = NULL, y = "Importance Score") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

---

# 14. Best Model Deep Dive

```{r best-model-id}
best_algo <- results_df$Algorithm[which.max(results_df$Macro_F1)]
best_f1   <- max(results_df$Macro_F1)
best_acc  <- results_df$Test_Accuracy[which.max(results_df$Macro_F1)]
cat("Best model:", best_algo, "\n")
cat("Macro F1:", best_f1, "| Test Accuracy:", best_acc, "\n")
cat("Improvement over baseline (33.3%):",
    round((best_acc - 0.333) * 100, 1), "percentage points\n")
```

## 14.1 Detailed Classification Report

```{r best-model-report}
best_cm <- switch(best_algo,
  "Logistic Regression" = lr_cm,
  "KNN"                 = knn_cm,
  "Random Forest"       = rf_cm,
  "GBM"                 = gbm_cm
)

detail_df <- as.data.frame(best_cm$byClass) %>%
  rownames_to_column("Class") %>%
  mutate(Class = str_remove(Class, "Class: ")) %>%
  rename(Balanced_Accuracy = `Balanced Accuracy`) %>%
  dplyr::select(Class, Precision, Recall, F1, Specificity, Balanced_Accuracy) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4)))

kable(detail_df, caption = paste("Detailed Classification Report —", best_algo)) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = TRUE) %>%
  column_spec(4, bold = TRUE, background = "#eaf4fb")
```

## 14.2 Prediction Confidence Distribution

```{r confidence-dist, fig.height=5}
best_probs <- switch(best_algo,
  "Logistic Regression" = lr_probs,
  "KNN"                 = knn_probs,
  "Random Forest"       = rf_probs,
  "GBM"                 = gbm_probs
)
best_pred <- switch(best_algo,
  "Logistic Regression" = lr_pred,
  "KNN"                 = knn_pred,
  "Random Forest"       = rf_pred,
  "GBM"                 = gbm_pred
)

conf_df <- best_probs %>%
  mutate(Predicted = best_pred,
         Actual    = test_df$Efficiency_Class,
         Correct   = Predicted == Actual,
         Max_Prob  = pmax(Low, Medium, High))

ggplot(conf_df, aes(x = Max_Prob, fill = Correct)) +
  geom_histogram(bins = 30, color = "white", alpha = 0.85) +
  scale_fill_manual(values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
                    labels = c("TRUE" = "Correct", "FALSE" = "Wrong")) +
  facet_wrap(~ Actual, ncol = 3) +
  labs(title = paste(best_algo, "— Prediction Confidence by Actual Class"),
       subtitle = "Green = correct | Red = wrong | X-axis = max predicted probability",
       x = "Max Class Probability", y = "Count", fill = "Prediction") +
  theme_minimal(base_size = 12)
```

---

# 15. Final Summary

```{r final-summary, echo=FALSE}
summary_tbl <- data.frame(
  Item = c(
    "Dataset",
    "Original target (rejected)",
    "New target",
    "Target construction",
    "Leakage prevention",
    "Classes",
    "Class balance",
    "Train / Test split",
    "Cross-validation",
    "Feature scaling",
    "Algorithms tried",
    "Primary metric",
    "Best algorithm",
    "Best Macro F1",
    "Best Test Accuracy",
    "Improvement over baseline"
  ),
  Detail = c(
    "Ship_Performance_Dataset.csv — 2,736 rows × 18 columns",
    "Maintenance_Status — randomly assigned, ANOVA F < 2 across all features",
    "Efficiency_Class (Low / Medium / High)",
    "ntile(Efficiency_nm_per_kWh, 3) — equal-frequency binning into 3 groups",
    "Efficiency_nm_per_kWh excluded from feature set",
    "3 ordered classes: Low < Medium < High",
    "Exactly 33.3% each — perfectly balanced by ntile construction",
    "80% train / 20% test — stratified by class",
    "5-fold stratified CV on training set",
    "Z-score (centre + scale) inside caret pipeline",
    "Logistic Regression, KNN, Random Forest, GBM",
    "Macro F1-Score (equal weight across all 3 classes)",
    best_algo,
    as.character(best_f1),
    as.character(best_acc),
    paste0(round((best_acc - 0.333) * 100, 1), " percentage points above 33.3% baseline")
  )
)

kable(summary_tbl, caption = "Classification Pipeline — Final Summary") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = TRUE) %>%
  row_spec(c(13,14,15,16), bold = TRUE, background = "#d5f5e3")
```

---

*End of Classification Report*
