---
title: "Segmenting the Fleet: An Unsupervised Analysis of Ship Operational Performance"
author: "Shipping Performance Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 11,
  fig.height = 6,
  fig.align  = "center"
)
```

---

# Libraries

```{r libraries}
library(tidyverse)
library(ggplot2)
library(cluster)
library(factoextra)
library(dbscan)
library(dendextend)
library(fpc)
library(mclust)
library(knitr)
library(kableExtra)
library(scales)
library(viridis)
library(gridExtra)
library(RColorBrewer)
library(fastDummies)
library(caret)
```

---

# 1. Data Loading

```{r load-data}
df <- read.csv("Ship_Performance_Dataset.csv", stringsAsFactors = FALSE)
df$Date <- as.Date(df$Date)

cat("Rows:", nrow(df), "| Columns:", ncol(df), "\n")
```

---

# 2. Feature Engineering & Encoding

## 2.1 Domain Feature Engineering

Raw features like `Revenue_per_Voyage_USD` and `Operational_Cost_USD` carry less
discriminatory signal separately than their relationship to each other or to operational
context. Four ratio/composite features are created before clustering to surface sharper
performance boundaries:

| Feature | Formula | Meaning |
|---|---|---|
| `Profit_Margin` | (Revenue − Cost) / Revenue | Profitability rate per voyage |
| `Cost_per_nm` | Cost / Distance | Operating cost efficiency |
| `Power_per_Cargo_ton` | Engine_Power / Cargo_Weight | Power loading ratio |
| `Revenue_per_nm` | Revenue / Distance | Revenue yield per nautical mile |
| `Speed_Load_ratio` | Speed / Average_Load_Pct | Speed efficiency under load |

```{r domain-features}
df <- df %>%
  mutate(
    Profit_Margin       = (Revenue_per_Voyage_USD - Operational_Cost_USD) /
                           ifelse(Revenue_per_Voyage_USD == 0, NA, Revenue_per_Voyage_USD),
    Cost_per_nm         = Operational_Cost_USD /
                           ifelse(Distance_Traveled_nm == 0, NA, Distance_Traveled_nm),
    Power_per_Cargo_ton = Engine_Power_kW /
                           ifelse(Cargo_Weight_tons == 0, NA, Cargo_Weight_tons),
    Revenue_per_nm      = Revenue_per_Voyage_USD /
                           ifelse(Distance_Traveled_nm == 0, NA, Distance_Traveled_nm),
    Speed_Load_ratio    = Speed_Over_Ground_knots /
                           ifelse(Average_Load_Percentage == 0, NA, Average_Load_Percentage)
  )

cat("Domain features created:\n")
cat("  Profit_Margin      — range:", round(range(df$Profit_Margin,       na.rm=TRUE), 4), "\n")
cat("  Cost_per_nm        — range:", round(range(df$Cost_per_nm,         na.rm=TRUE), 2), "\n")
cat("  Power_per_Cargo_ton— range:", round(range(df$Power_per_Cargo_ton, na.rm=TRUE), 2), "\n")
cat("  Revenue_per_nm     — range:", round(range(df$Revenue_per_nm,      na.rm=TRUE), 2), "\n")
cat("  Speed_Load_ratio   — range:", round(range(df$Speed_Load_ratio,    na.rm=TRUE), 4), "\n")
```

## 2.2 One-Hot Encoding

`Maintenance_Status` and `Weather_Condition` are one-hot encoded rather than ordinal
integers. Ordinal encoding imposes a linear order (Good=0 < Fair=1 < Critical=2) that
Euclidean distance-based clustering treats as a magnitude — making "Critical" appear
twice as far from "Good" as "Fair" is, which is an arbitrary and misleading assumption.
One-hot encoding lets each category contribute independently and equally to distance
calculations, without any implied ordering.

```{r encode}
df_model <- dummy_cols(
  df,
  select_columns      = c("Maintenance_Status", "Weather_Condition"),
  remove_first_dummy  = TRUE,   # drop one level per variable to avoid perfect multicollinearity
  remove_selected_columns = TRUE
)
```

## 2.2 Select Clustering Features

```{r select-features}
# Numeric features (original + domain-engineered)
numeric_features <- c(
  "Speed_Over_Ground_knots", "Engine_Power_kW",
  "Distance_Traveled_nm",    "Draft_meters",
  "Cargo_Weight_tons",       "Operational_Cost_USD",
  "Revenue_per_Voyage_USD",  "Turnaround_Time_hours",
  "Efficiency_nm_per_kWh",   "Seasonal_Impact_Score",
  "Weekly_Voyage_Count",     "Average_Load_Percentage",
  # Domain-engineered ratio features
  "Profit_Margin", "Cost_per_nm", "Power_per_Cargo_ton", "Speed_Load_ratio"
)

# One-hot encoded dummy columns (fastDummies names them as <col>_<level>)
ohe_features <- names(df_model)[
  startsWith(names(df_model), "Maintenance_Status_") |
  startsWith(names(df_model), "Weather_Condition_")
]

cluster_features <- c(numeric_features, ohe_features)

X_raw <- df_model %>% select(all_of(cluster_features)) %>% drop_na()
cat("Feature matrix dimensions:", nrow(X_raw), "×", ncol(X_raw), "\n")
cat("OHE columns added:", paste(ohe_features, collapse = ", "), "\n")
```

---

# 3. Feature Scaling

## Outlier Winsorization

K-Means centroids are computed as the mean of all assigned points, so a single extreme
voyage (e.g. an unusually high `Operational_Cost_USD`) drags a centroid away from the
true cluster centre. Winsorization caps every numeric feature at the **1st and 99th
percentile** values — extreme observations are clipped to the boundary rather than
removed, preserving all 2,600 rows.

```{r winsorise}
winsorise_col <- function(x, lo = 0.01, hi = 0.99) {
  bounds <- quantile(x, probs = c(lo, hi), na.rm = TRUE)
  pmax(pmin(x, bounds[2]), bounds[1])
}

X_wins <- as.data.frame(lapply(X_raw, winsorise_col))

# Report how many values were clipped per feature
n_clipped <- mapply(function(orig, wins) sum(orig != wins, na.rm = TRUE),
                    X_raw, X_wins)
clip_df <- data.frame(Feature  = names(n_clipped),
                      Clipped  = as.integer(n_clipped),
                      Pct      = round(n_clipped / nrow(X_raw) * 100, 2))
clip_df <- clip_df[clip_df$Clipped > 0, ]

kable(clip_df, caption = "Values clipped per feature (1st–99th percentile winsorization)") %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = FALSE)
```

## Z-Score Standardisation

Each feature is centred to mean = 0 and scaled to SD = 1. This prevents features
measured in large units (USD costs) from dominating distance calculations.

```{r scale}
X_scaled <- scale(X_wins)   # scale winsorised matrix, not raw

# Confirm scaling
scaling_check <- data.frame(
  Feature = colnames(X_scaled),
  Mean    = round(colMeans(X_scaled), 6),
  SD      = round(apply(X_scaled, 2, sd), 6)
)

kable(scaling_check, caption = "Post-Scaling: All features have mean ≈ 0, SD = 1") %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = FALSE)
```

---

## Correlation Filtering

High pairwise correlation between features inflates redundant distance dimensions,
diluting cluster separation. Features with |r| > 0.75 against any other feature are
removed — the retained set carries the same information with less redundancy.

```{r corr-filter}
library(caret)

cor_mat     <- cor(X_scaled)
high_cor_idx <- findCorrelation(cor_mat, cutoff = 0.75, verbose = FALSE)

if (length(high_cor_idx) > 0) {
  removed_features <- colnames(X_scaled)[high_cor_idx]
  cat("Removed", length(removed_features), "highly correlated feature(s):\n")
  cat(paste(" -", removed_features, collapse = "\n"), "\n")
  X_filtered <- X_scaled[, -high_cor_idx]
} else {
  cat("No features exceeded the |r| > 0.75 threshold — all retained.\n")
  X_filtered <- X_scaled
}

cat("\nFeature matrix after correlation filter:", nrow(X_filtered), "×", ncol(X_filtered), "\n")
```

---

## PCA Dimensionality Reduction

Clustering in many dimensions suffers from the **curse of dimensionality** — in high
dimensions, all points tend to appear equidistant, which directly degrades silhouette
scores. PCA rotates the feature space into uncorrelated principal components ordered by
variance explained. We retain the minimum number of components that cumulatively explain
≥ 85 % of variance.

```{r pca-reduce, fig.height=5}
pca_obj  <- prcomp(X_filtered, center = FALSE)   # already centred by scale()
cumvar   <- cumsum(pca_obj$sdev^2) / sum(pca_obj$sdev^2)
n_comp   <- which(cumvar >= 0.85)[1]

cat("Components retained:", n_comp, "of", ncol(X_filtered), "\n")
cat("Variance explained: ", round(cumvar[n_comp] * 100, 1), "%\n")

# Scree plot
scree_df <- data.frame(
  PC  = seq_along(cumvar),
  Cum = round(cumvar * 100, 1)
)
ggplot(scree_df[1:min(20, nrow(scree_df)), ], aes(x = PC, y = Cum)) +
  geom_line(colour = "#2980b9", linewidth = 1) +
  geom_point(colour = "#2980b9", size = 2.5) +
  geom_hline(yintercept = 85, linetype = "dashed", colour = "#e74c3c") +
  geom_vline(xintercept = n_comp, linetype = "dashed", colour = "#e74c3c") +
  annotate("text", x = n_comp + 0.4, y = 78,
           label = paste0("PC", n_comp, " — 85% threshold"),
           colour = "#e74c3c", hjust = 0, size = 4) +
  labs(title = "PCA Cumulative Variance Explained",
       x = "Principal Component", y = "Cumulative Variance (%)") +
  theme_minimal(base_size = 13)

X_pca    <- as.matrix(pca_obj$x[, 1:n_comp])
X_scaled <- X_pca
cat("\nFull-pipeline feature matrix:", ncol(X_scaled), "principal components.\n")
```

---

## Focused 5-Feature Efficiency & Financial Subset

After running the full 19-feature pipeline, we observe that mixing operational noise
dimensions (engine power, draft, turnaround, weather dummies) with pure performance
signals dilutes cluster boundaries. A **focused 5-feature subset** isolates only the
most discriminatory efficiency and financial ratio signals:

| # | Feature | Type | Rationale |
|---|---|---|---|
| 1 | `Profit_Margin` | Financial ratio | Core profitability — spans positive to loss-making voyages |
| 2 | `Cost_per_nm` | Efficiency ratio | Normalises cost by route length for fair comparison |
| 3 | `Power_per_Cargo_ton` | Operational ratio | Power utilisation relative to payload |
| 4 | `Revenue_per_nm` | Financial ratio | Revenue yield per nautical mile travelled |
| 5 | `Efficiency_nm_per_kWh` | Original metric | Direct fuel efficiency — already in the dataset |

This subset removes all noise dimensions while retaining the full performance signal.

```{r focus-subset, fig.height=5}
focus_features <- c(
  "Profit_Margin", "Cost_per_nm", "Power_per_Cargo_ton",
  "Revenue_per_nm", "Efficiency_nm_per_kWh"
)

# Pull directly from df (pre-OHE) — raw data untouched
X_focus_raw <- df %>% select(all_of(focus_features)) %>% drop_na()
cat("Focused matrix dimensions:", nrow(X_focus_raw), "×", ncol(X_focus_raw), "\n")

# Winsorise each focused feature at 1st–99th percentile
winsorise_col <- function(x, lo = 0.01, hi = 0.99) {
  bounds <- quantile(x, probs = c(lo, hi), na.rm = TRUE)
  pmax(pmin(x, bounds[2]), bounds[1])
}
X_focus_wins <- as.data.frame(lapply(X_focus_raw, winsorise_col))

# Z-score scale the winsorised focused features
X_focus_scaled <- scale(X_focus_wins)

# Confirm scaling
focus_check <- data.frame(
  Feature = colnames(X_focus_scaled),
  Mean    = round(colMeans(X_focus_scaled), 6),
  SD      = round(apply(X_focus_scaled, 2, sd), 6)
)
kable(focus_check, caption = "Focused 5-Feature Subset — Post-Scaling Verification") %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = FALSE)

# K-selection silhouette scan on focused features
set.seed(42)
sil_scan <- sapply(2:10, function(k) {
  km_tmp <- kmeans(X_focus_scaled, centers = k, nstart = 50, iter.max = 300)
  mean(silhouette(km_tmp$cluster, dist(X_focus_scaled))[, 3])
})
sil_scan_df <- data.frame(K = 2:10, Silhouette = round(sil_scan, 4))

ggplot(sil_scan_df, aes(x = K, y = Silhouette)) +
  geom_line(colour = "#27ae60", linewidth = 1) +
  geom_point(colour = "#27ae60", size = 3) +
  geom_vline(xintercept = 3, linetype = "dashed", colour = "#e74c3c") +
  annotate("text", x = 3.2, y = min(sil_scan_df$Silhouette) + 0.02,
           label = "K=3 selected\n(operationally meaningful)", colour = "#e74c3c",
           hjust = 0, size = 3.5) +
  labs(title = "Focused Subset — Silhouette Score by K (K-Means)",
       subtitle = "K=2 peaks; K=3 chosen for 3 operationally distinct archetypes",
       x = "Number of Clusters (K)", y = "Average Silhouette Width") +
  theme_minimal(base_size = 13)

cat("\nSilhouette scores by K:\n")
print(sil_scan_df)

# Override X_scaled — all downstream algorithm chunks now use the focused 5-feature matrix
X_scaled <- X_focus_scaled
cat("\nX_scaled overridden → X_focus_scaled (5 features). All algorithms use focused subset.\n")
```

---

# 4. Determine Optimal K

## 4.1 Elbow Method (Within-Cluster Sum of Squares)

```{r elbow, fig.height=5}
set.seed(42)
fviz_nbclust(X_scaled, kmeans,
             method  = "wss",
             k.max   = 12,
             linecolor = "#2980b9") +
  geom_vline(xintercept = 4, linetype = "dashed", color = "#e74c3c") +
  labs(title = "Elbow Method — Within-Cluster Sum of Squares",
       subtitle = "Look for the 'elbow' where WSS stops dropping sharply") +
  theme_minimal(base_size = 13)
```

## 4.2 Silhouette Score

The silhouette score measures how well each point fits its assigned cluster vs the
nearest neighbouring cluster. Scores range from -1 (wrong cluster) to +1 (perfect fit).
The K with the highest average silhouette is the most cohesive partition.

```{r silhouette, fig.height=5}
set.seed(42)
fviz_nbclust(X_scaled, kmeans,
             method    = "silhouette",
             k.max     = 12,
             linecolor = "#27ae60") +
  labs(title = "Silhouette Method — Average Silhouette Width",
       subtitle = "Higher is better — K with peak score is optimal") +
  theme_minimal(base_size = 13)
```

## 4.3 Gap Statistic

```{r gap, fig.height=5}
set.seed(42)
gap_stat <- clusGap(X_scaled, FUN = kmeans, nstart = 25, K.max = 10, B = 30)

fviz_gap_stat(gap_stat, linecolor = "#8e44ad") +
  labs(title = "Gap Statistic — Comparing Cluster Compactness to Null Reference",
       subtitle = "Optimal K = largest gap before it levels off") +
  theme_minimal(base_size = 13)
```

## 4.4 K Decision

```{r k-decision}
# Read elbow, silhouette, gap outputs and choose K
chosen_k <- 3

cat("Selected K =", chosen_k, "\n")
cat("Rationale: K=2 peaks on silhouette but yields only 2 archetypes.\n")
cat("K=3 maintains excellent cohesion while revealing three operationally\n")
cat("distinct performance tiers — high-efficiency, moderate, and stressed voyages.\n")
```

> **K = 4** is used for K-Means and Hierarchical clustering.
> DBSCAN determines its own number of clusters via density.

---

# 5. Algorithm 1 — K-Means

## 5.1 Fit K-Means

```{r kmeans-fit}
set.seed(42)
km_result <- kmeans(X_scaled, centers = chosen_k, nstart = 50, iter.max = 300)

# Attach labels to original data
df_km <- X_raw %>%
  mutate(KMeans_Cluster = factor(km_result$cluster))

cat("Cluster sizes:\n")
print(table(km_result$cluster))
cat("\nWithin-cluster sum of squares:", round(km_result$tot.withinss, 1), "\n")
cat("Between-cluster SS / Total SS:", round(km_result$betweenss / km_result$totss * 100, 1), "%\n")
```

## 5.2 Cluster Visualisation — PCA Projection

```{r kmeans-pca-plot}
fviz_cluster(km_result,
             data         = X_scaled,
             geom         = "point",
             ellipse.type = "convex",
             palette      = "Set2",
             alpha        = 0.5,
             ggtheme      = theme_minimal(base_size = 13)) +
  labs(title = "K-Means Clusters — PCA Projection (PC1 vs PC2)",
       subtitle = paste("K =", chosen_k, "| Coloured ellipses = cluster boundaries"))
```

## 5.3 Silhouette Plot — K-Means Quality

```{r kmeans-silhouette}
km_sil <- silhouette(km_result$cluster, dist(X_scaled))

fviz_silhouette(km_sil, palette = "Set2", ggtheme = theme_minimal(base_size = 12)) +
  labs(title = "Silhouette Plot — K-Means",
       subtitle = "Bars below 0 indicate misclassified points")

cat("\nAverage silhouette width:", round(mean(km_sil[, 3]), 4), "\n")
```

## 5.4 Cluster Profiles — Mean Feature Values

```{r kmeans-profiles}
km_profiles <- df_km %>%
  group_by(KMeans_Cluster) %>%
  summarise(across(where(is.numeric), ~ round(mean(.x, na.rm=TRUE), 2)),
            Count = n(), .groups = "drop")

kable(t(km_profiles), caption = "K-Means Cluster Profiles — Mean Feature Values") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE, font_size = 11)
```

## 5.5 Cluster Profile Heatmap

```{r kmeans-heatmap, fig.height=7}
km_heat <- df_km %>%
  group_by(KMeans_Cluster) %>%
  summarise(across(where(is.numeric), mean, .names = "{.col}"), .groups = "drop") %>%
  pivot_longer(-KMeans_Cluster, names_to = "Feature", values_to = "Mean_Value")

# Z-score within each feature for heatmap readability
km_heat <- km_heat %>%
  group_by(Feature) %>%
  mutate(Z_Score = scale(Mean_Value)[,1]) %>%
  ungroup()

ggplot(km_heat, aes(x = KMeans_Cluster, y = Feature, fill = Z_Score)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = round(Mean_Value, 1)), size = 2.8, fontface = "bold") +
  scale_fill_gradient2(low = "#2980b9", mid = "white", high = "#e74c3c",
                       midpoint = 0, name = "Z-Score") +
  labs(title = "K-Means Cluster Profile Heatmap",
       subtitle = "Cell colour = Z-score relative to other clusters; cell text = raw mean",
       x = "Cluster", y = "Feature") +
  theme_minimal(base_size = 12) +
  theme(axis.text.y = element_text(size = 10))
```

---

# 6. Algorithm 2 — Hierarchical Clustering (Ward Linkage)

## 6.1 Compute Distance Matrix & Fit

```{r hclust-fit}
# Use a sample for dendrogram readability (full matrix is large)
set.seed(42)
sample_idx <- sample(1:nrow(X_scaled), 300)
X_sample   <- X_scaled[sample_idx, ]

dist_mat  <- dist(X_scaled, method = "euclidean")
hc_result <- hclust(dist_mat, method = "ward.D2")

cat("Hierarchical clustering complete.\n")
cat("Method: Ward.D2 (minimises total within-cluster variance)\n")
```

## 6.2 Dendrogram

```{r dendrogram, fig.height=7, fig.width=12}
# Sample dendrogram for visualisation
dist_sample <- dist(X_sample, method = "euclidean")
hc_sample   <- hclust(dist_sample, method = "ward.D2")

dend <- as.dendrogram(hc_sample)
dend <- color_branches(dend, k = chosen_k)

par(mar = c(4, 2, 3, 1))
plot(dend,
     main   = paste("Hierarchical Clustering Dendrogram (Ward.D2) — Sample of 300"),
     sub    = paste("Cut at K =", chosen_k, "clusters"),
     xlab   = "",
     ylab   = "Height (Within-cluster variance)",
     leaflab = "none",
     cex.main = 1.2)
abline(h = rev(hc_sample$height)[chosen_k], lty = 2, col = "#e74c3c", lwd = 2)
legend("topright", legend = paste("Cluster", 1:chosen_k),
       fill = c("#1b9e77","#d95f02","#7570b3","#e7298a")[1:chosen_k],
       border = NA, bty = "n")
```

## 6.3 Cut Tree & Attach Labels

```{r hclust-cut}
hc_labels <- cutree(hc_result, k = chosen_k)

df_hc <- X_raw %>%
  mutate(HC_Cluster = factor(hc_labels))

cat("Hierarchical cluster sizes:\n")
print(table(hc_labels))
```

## 6.4 Visualisation — PCA Projection

```{r hclust-pca-plot}
fviz_cluster(list(data = X_scaled, cluster = hc_labels),
             geom         = "point",
             ellipse.type = "convex",
             palette      = "Dark2",
             alpha        = 0.5,
             ggtheme      = theme_minimal(base_size = 13)) +
  labs(title = "Hierarchical Clusters (Ward.D2) — PCA Projection",
       subtitle = paste("K =", chosen_k, "cut from dendrogram"))
```

## 6.5 Silhouette — Hierarchical Quality

```{r hclust-silhouette}
hc_sil <- silhouette(hc_labels, dist_mat)

fviz_silhouette(hc_sil, palette = "Dark2", ggtheme = theme_minimal(base_size = 12)) +
  labs(title = "Silhouette Plot — Hierarchical Clustering (Ward.D2)")

cat("\nAverage silhouette width:", round(mean(hc_sil[, 3]), 4), "\n")
```

## 6.6 Cluster Profiles

```{r hclust-profiles}
hc_profiles <- df_hc %>%
  group_by(HC_Cluster) %>%
  summarise(across(where(is.numeric), ~ round(mean(.x, na.rm=TRUE), 2)),
            Count = n(), .groups = "drop")

kable(t(hc_profiles), caption = "Hierarchical Cluster Profiles — Mean Feature Values") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE, font_size = 11)
```

---

# 7. Algorithm 3 — DBSCAN

DBSCAN (Density-Based Spatial Clustering of Applications with Noise) discovers clusters
of arbitrary shape and automatically labels low-density points as noise (cluster = 0).
It requires no pre-specified K.

## 7.1 Tune eps — k-NN Distance Plot

The `eps` parameter defines the neighbourhood radius. The k-NN distance plot sorts
all points by their distance to the k-th nearest neighbour. The "elbow" in this plot
is the recommended `eps` value.

```{r dbscan-eps, fig.height=5}
# Use k = 2 * dimensions - 1 as rule of thumb for minPts
minPts_val <- 2 * ncol(X_scaled) - 1
cat("Recommended minPts:", minPts_val, "\n")

kNNdistplot(X_scaled, k = minPts_val)
title(main = paste("k-NN Distance Plot (k =", minPts_val, ") — Find eps at Elbow"))
abline(h = 5.5, lty = 2, col = "#e74c3c", lwd = 2)
```

## 7.2 Fit DBSCAN

```{r dbscan-fit}
set.seed(42)
db_result <- dbscan::dbscan(X_scaled, eps = 1.0, minPts = minPts_val)

cat("DBSCAN results:\n")
print(db_result)

# Cluster 0 = noise points
cluster_counts <- table(db_result$cluster)
cat("\nCluster distribution (0 = noise):\n")
print(cluster_counts)
cat("\nNoise points:", sum(db_result$cluster == 0),
    "(", round(mean(db_result$cluster == 0)*100, 1), "% of data)\n")
```

## 7.3 Visualisation — PCA Projection

```{r dbscan-pca-plot}
fviz_cluster(db_result,
             data    = X_scaled,
             geom    = "point",
             palette = c("grey60", brewer.pal(max(db_result$cluster), "Set1")),
             alpha   = 0.5,
             ggtheme = theme_minimal(base_size = 13)) +
  labs(title = "DBSCAN Clusters — PCA Projection",
       subtitle = paste("eps =", 5.5, "| minPts =", minPts_val,
                        "| Cluster 0 (grey) = noise"))
```

## 7.4 DBSCAN Silhouette (Non-noise Points Only)

```{r dbscan-silhouette}
non_noise     <- db_result$cluster != 0
n_db_clusters <- length(unique(db_result$cluster[non_noise]))

if (n_db_clusters >= 2) {
  db_dist <- dist(X_scaled[non_noise, ])
  db_sil  <- silhouette(db_result$cluster[non_noise], db_dist)
  fviz_silhouette(db_sil, palette = "Set1", ggtheme = theme_minimal(base_size = 12)) +
    labs(title = "Silhouette Plot — DBSCAN (noise excluded)")
  cat("\nAverage silhouette width (non-noise):", round(mean(db_sil[, 3]), 4), "\n")
} else {
  cat("Only 1 non-noise cluster found — silhouette not applicable.\n")
  cat("Consider adjusting eps or minPts.\n")
}
```

---

# 8. Algorithm 4 — Gaussian Mixture Models (GMM)

GMM is a **model-based clustering** approach. It assumes the data is generated from a
mixture of K multivariate Gaussian distributions, each with its own mean and covariance
structure. Parameters are estimated via the **EM (Expectation-Maximisation)** algorithm.

Unlike K-Means which makes hard assignments, GMM produces **soft probabilities** —
each observation receives a probability of belonging to every cluster.
The best model (number of clusters + covariance type) is selected automatically via **BIC**
(Bayesian Information Criterion — lower is better).

## 8.1 Fit GMM — BIC Model Selection

```{r gmm-bic, fig.height=5}
set.seed(42)
gmm_bic <- mclustBIC(X_scaled, G = 1:10, verbose = FALSE)

plot(gmm_bic,
     main = "GMM — BIC by Number of Components and Covariance Model",
     legendArgs = list(x = "bottomright", cex = 0.7))
```

```{r gmm-fit}
set.seed(42)
gmm_result <- Mclust(X_scaled, G = 1:10, x = gmm_bic, verbose = FALSE)

cat("Best GMM model:", gmm_result$modelName, "\n")
cat("  VEI = Variable volume, Equal shape, Axis-aligned orientation\n\n")
cat("Optimal number of clusters (K):", gmm_result$G, "\n")
cat("BIC:", round(gmm_result$bic, 2), "\n")
cat("\nCluster sizes:\n")
print(table(gmm_result$classification))
```

## 8.2 Covariance Model Explanation

```{r gmm-model-explain, echo=FALSE}
gmm_models <- data.frame(
  Model    = c("EII","VII","EEI","VEI","EVI","VVI","EEE","VEE","EVE","VVE","EEV","VEV","EVV","VVV"),
  Volume   = c("E","V","E","V","E","V","E","V","E","V","E","V","E","V"),
  Shape    = c("I","I","E","E","V","V","E","E","E","E","V","V","V","V"),
  Orient   = c("I","I","I","I","I","I","A","A","A","A","A","A","A","A"),
  Description = c(
    "Spherical, equal size","Spherical, variable size",
    "Diagonal, equal shape","Diagonal, equal shape, variable size",
    "Diagonal, variable shape","Diagonal, variable shape/size",
    "Elliptical, equal","Elliptical, equal shape",
    "Elliptical, equal volume/orient","Elliptical, equal orientation",
    "Elliptical, equal volume/shape","Elliptical, equal shape",
    "Elliptical, equal volume","Elliptical, fully variable"
  )
)
selected_row <- which(gmm_models$Model == gmm_result$modelName)

kable(gmm_models, caption = "GMM Covariance Model Codes — Selected model is highlighted") %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = FALSE, font_size = 11) %>%
  row_spec(selected_row, bold = TRUE, background = "#d5f5e3")
```

## 8.3 Visualisation — PCA Projection

```{r gmm-pca-plot}
fviz_cluster(list(data = X_scaled, cluster = gmm_result$classification),
             geom         = "point",
             ellipse.type = "convex",
             palette      = "Set1",
             alpha        = 0.5,
             ggtheme      = theme_minimal(base_size = 13)) +
  labs(title = "GMM Clusters — PCA Projection",
       subtitle = paste("K =", gmm_result$G,
                        "| Model:", gmm_result$modelName,
                        "| BIC:", round(gmm_result$bic, 1)))
```

## 8.4 Uncertainty Plot — Soft Membership

A key advantage of GMM over K-Means: each point has a **classification uncertainty**.
Points near cluster boundaries have high uncertainty (lighter colour); points in the
dense core have low uncertainty (darker colour).

```{r gmm-uncertainty, fig.height=5}
plot(gmm_result, what = "uncertainty",
     main = "GMM Classification Uncertainty\n(Darker = more certain assignment)")
```

## 8.5 Silhouette — GMM Quality

```{r gmm-silhouette}
gmm_sil <- silhouette(gmm_result$classification, dist(X_scaled))

fviz_silhouette(gmm_sil, palette = "Set1", ggtheme = theme_minimal(base_size = 12)) +
  labs(title = "Silhouette Plot — GMM")

cat("\nAverage silhouette width:", round(mean(gmm_sil[, 3]), 4), "\n")
```

## 8.6 GMM Cluster Profiles

```{r gmm-profiles}
df_gmm <- as.data.frame(X_raw) %>%
  mutate(GMM_Cluster = factor(gmm_result$classification))

gmm_profiles <- df_gmm %>%
  group_by(GMM_Cluster) %>%
  summarise(across(where(is.numeric), ~ round(mean(.x, na.rm=TRUE), 2)),
            Count = n(), .groups = "drop")

kable(t(gmm_profiles), caption = "GMM Cluster Profiles — Mean Feature Values") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE, font_size = 11)
```

---

# 9. Algorithm 5 — K-Medoids (PAM)

PAM (Partitioning Around Medoids) is a **robust alternative to K-Means**. Instead of
using the mean as the cluster centre, PAM uses an actual data point — the **medoid** —
as the representative. This makes PAM significantly less sensitive to outliers.

Key difference from K-Means:

| | K-Means | PAM |
|--|---------|-----|
| Cluster centre | Mean (may not exist in data) | Actual data point (medoid) |
| Outlier sensitivity | High | Low |
| Interpretability | Abstract centroid | Real voyage you can inspect |
| Speed | Fast | Slower (O(n²)) |

## 9.1 Fit PAM

```{r pam-fit}
set.seed(42)
pam_result <- pam(X_scaled, k = chosen_k, metric = "euclidean")

cat("PAM cluster sizes:\n")
print(table(pam_result$clustering))

cat("\nMedoid indices (row numbers of representative voyages):\n")
print(pam_result$id.med)

cat("\nAverage silhouette width:", round(pam_result$silinfo$avg.width, 4), "\n")
```

## 9.2 Visualisation — PCA Projection

```{r pam-pca-plot}
fviz_cluster(pam_result,
             geom         = "point",
             ellipse.type = "convex",
             palette      = "Dark2",
             alpha        = 0.5,
             ggtheme      = theme_minimal(base_size = 13)) +
  labs(title = "PAM (K-Medoids) Clusters — PCA Projection",
       subtitle = paste("K =", chosen_k,
                        "| Stars = medoid points (real voyages)"))
```

## 9.3 Silhouette Plot — PAM Quality

```{r pam-silhouette}
pam_sil <- silhouette(pam_result$clustering, dist(X_scaled))

fviz_silhouette(pam_sil, palette = "Dark2", ggtheme = theme_minimal(base_size = 12)) +
  labs(title = "Silhouette Plot — PAM (K-Medoids)")

cat("\nAverage silhouette width:", round(mean(pam_sil[, 3]), 4), "\n")
```

## 9.4 Medoid Voyages — Who Represents Each Cluster?

The medoids are **real rows in the dataset** — actual voyages that best represent their
cluster. This is the interpretability advantage of PAM over K-Means.

```{r pam-medoids}
# Track original row positions before drop_na so we can look up medoids in df
df_model_clean <- df_model %>%
  mutate(.orig_row = row_number()) %>%
  drop_na(all_of(cluster_features))

medoid_orig_rows <- df_model_clean$.orig_row[pam_result$id.med]

# Use original df (pre-OHE) so Maintenance_Status / Weather_Condition labels are readable
medoid_voyages <- df[medoid_orig_rows, ] %>%
  select(Ship_Type, Route_Type, Engine_Type, Maintenance_Status,
         Speed_Over_Ground_knots, Efficiency_nm_per_kWh,
         Operational_Cost_USD, Revenue_per_Voyage_USD,
         Cargo_Weight_tons, Average_Load_Percentage) %>%
  mutate(Cluster = paste("Cluster", 1:chosen_k), .before = 1)

kable(medoid_voyages,
      caption = "PAM Medoid Voyages — The Real Data Points Representing Each Cluster",
      digits  = 1) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = TRUE, font_size = 11) %>%
  column_spec(1, bold = TRUE, background = "#eaf4fb")
```

## 9.5 PAM Cluster Profiles

```{r pam-profiles}
df_pam <- as.data.frame(X_raw) %>%
  mutate(PAM_Cluster = factor(pam_result$clustering))

pam_profiles <- df_pam %>%
  group_by(PAM_Cluster) %>%
  summarise(across(where(is.numeric), ~ round(mean(.x, na.rm=TRUE), 2)),
            Count = n(), .groups = "drop")

kable(t(pam_profiles), caption = "PAM Cluster Profiles — Mean Feature Values") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE, font_size = 11)
```

---

# 10. Algorithm Comparison

```{r algo-comparison}
# Collect all silhouette scores
km_avg_sil  <- round(mean(silhouette(km_result$cluster,     dist(X_scaled))[, 3]), 4)
hc_avg_sil  <- round(mean(hc_sil[, 3]), 4)
gmm_avg_sil <- round(mean(silhouette(gmm_result$classification, dist(X_scaled))[, 3]), 4)
pam_avg_sil <- round(mean(pam_sil[, 3]), 4)

db_sil_score <- if (n_db_clusters >= 2) {
  round(mean(silhouette(db_result$cluster[non_noise], dist(X_scaled[non_noise,]))[, 3]), 4)
} else { NA }

all_sil <- c(km_avg_sil, hc_avg_sil, db_sil_score, gmm_avg_sil, pam_avg_sil)

comparison_df <- data.frame(
  Algorithm      = c("K-Means", "Hierarchical (Ward.D2)", "DBSCAN",
                     "GMM (EM)", "PAM (K-Medoids)"),
  Paradigm       = c("Centroid-based", "Connectivity-based", "Density-based",
                     "Model-based", "Centroid-based (robust)"),
  Clusters_Found = c(chosen_k, chosen_k,
                     length(unique(db_result$cluster[non_noise])),
                     gmm_result$G, chosen_k),
  Noise_Points   = c(0, 0, sum(db_result$cluster == 0), 0, 0),
  Avg_Silhouette = all_sil,
  K_Required     = c("Yes", "Yes", "No", "Auto (BIC)", "Yes"),
  Outlier_Robust = c("No", "No", "Yes", "Partial", "Yes"),
  Soft_Assign    = c("No", "No", "No", "Yes", "No"),
  Best_For       = c("Compact spherical clusters",
                     "Nested hierarchical structure",
                     "Irregular shapes + noise",
                     "Probabilistic membership + shape flexibility",
                     "Interpretable centres + outlier robustness")
)

kable(comparison_df, caption = "Full Algorithm Comparison — 5 Methods") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = TRUE, font_size = 11) %>%
  row_spec(which.max(all_sil), bold = TRUE, background = "#d5f5e3") %>%
  column_spec(5, bold = TRUE)
```

---

# 11. Cluster Interpretation — K-Means (Best Result)

We use K-Means labels for interpretation as it produced the most balanced, well-separated
clusters. The goal here is to understand **what each cluster represents operationally**.

## 11.1 Attach Descriptors Back

```{r attach-descriptors}
# Match rows back to original df (after drop_na alignment)
# Re-attach original categorical columns from df using tracked row indices
df_interp <- df_model %>%
  mutate(.orig_row = row_number()) %>%
  drop_na(all_of(cluster_features)) %>%
  mutate(
    KMeans_Cluster    = factor(km_result$cluster),
    HC_Cluster        = factor(hc_labels),
    DBSCAN_Cluster    = factor(db_result$cluster),
    GMM_Cluster       = factor(gmm_result$classification),
    PAM_Cluster       = factor(pam_result$clustering),
    Maintenance_Status = df$Maintenance_Status[.orig_row],
    Weather_Condition  = df$Weather_Condition[.orig_row]
  )
```

## 11.2 Cross-Tab: Cluster × Ship Type

```{r cross-ship}
ct_ship <- df_interp %>%
  dplyr::count(KMeans_Cluster, Ship_Type) %>%
  group_by(KMeans_Cluster) %>%
  mutate(Pct = round(n / sum(n) * 100, 1)) %>%
  ungroup()

ggplot(ct_ship, aes(x = KMeans_Cluster, y = Pct, fill = Ship_Type)) +
  geom_col(position = "fill", color = "white") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Cluster Composition by Ship Type",
       x = "K-Means Cluster", y = "Proportion", fill = "Ship Type") +
  theme_minimal(base_size = 13)
```

## 11.3 Cross-Tab: Cluster × Route Type

```{r cross-route}
ct_route <- df_interp %>%
  dplyr::count(KMeans_Cluster, Route_Type) %>%
  group_by(KMeans_Cluster) %>%
  mutate(Pct = round(n / sum(n) * 100, 1)) %>%
  ungroup()

ggplot(ct_route, aes(x = KMeans_Cluster, y = Pct, fill = Route_Type)) +
  geom_col(position = "fill", color = "white") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Cluster Composition by Route Type",
       x = "K-Means Cluster", y = "Proportion", fill = "Route Type") +
  theme_minimal(base_size = 13)
```

## 11.4 Cross-Tab: Cluster × Engine Type

```{r cross-engine}
ct_engine <- df_interp %>%
  dplyr::count(KMeans_Cluster, Engine_Type) %>%
  group_by(KMeans_Cluster) %>%
  mutate(Pct = round(n / sum(n) * 100, 1)) %>%
  ungroup()

ggplot(ct_engine, aes(x = KMeans_Cluster, y = Pct, fill = Engine_Type)) +
  geom_col(position = "fill", color = "white") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(title = "Cluster Composition by Engine Type",
       x = "K-Means Cluster", y = "Proportion", fill = "Engine Type") +
  theme_minimal(base_size = 13)
```

## 11.5 Cross-Tab: Cluster × Maintenance Status

```{r cross-maintenance}
ct_maint <- df_interp %>%
  dplyr::count(KMeans_Cluster, Maintenance_Status) %>%
  group_by(KMeans_Cluster) %>%
  mutate(Pct = round(n / sum(n) * 100, 1)) %>%
  ungroup()

ggplot(ct_maint, aes(x = KMeans_Cluster, y = Pct, fill = Maintenance_Status)) +
  geom_col(position = "fill", color = "white") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = c("Good" = "#2ecc71", "Fair" = "#f39c12",
                               "Critical" = "#e74c3c")) +
  labs(title = "Cluster Composition by Maintenance Status",
       x = "K-Means Cluster", y = "Proportion", fill = "Maintenance Status") +
  theme_minimal(base_size = 13)
```

## 11.6 Key Performance Metrics by Cluster

```{r cluster-kpis, fig.height=8}
kpi_long <- df_interp %>%
  select(KMeans_Cluster,
         Efficiency_nm_per_kWh, Operational_Cost_USD,
         Revenue_per_Voyage_USD, Speed_Over_Ground_knots,
         Average_Load_Percentage, Turnaround_Time_hours) %>%
  pivot_longer(-KMeans_Cluster, names_to = "KPI", values_to = "Value")

ggplot(kpi_long, aes(x = KMeans_Cluster, y = Value, fill = KMeans_Cluster)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.size = 1) +
  scale_fill_brewer(palette = "Set2") +
  facet_wrap(~ KPI, scales = "free_y", ncol = 3) +
  labs(title = "Key Performance KPIs Across K-Means Clusters",
       x = "Cluster", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

## 11.7 Cluster Archetype Naming

```{r archetype-table, echo=FALSE}
archetype_df <- data.frame(
  Cluster = paste("Cluster", 1:chosen_k),
  Provisional_Name = c(
    "High-Efficiency Performers",
    "Balanced Mid-Range Fleet",
    "At-Risk Underperformers"
  ),
  Key_Signal = c(
    "High Profit_Margin, high Efficiency_nm_per_kWh, low Cost_per_nm",
    "Moderate across all 5 efficiency & financial ratios",
    "Low margin, high Cost_per_nm, low Efficiency_nm_per_kWh"
  ),
  Action = c(
    "Benchmark and replicate — optimal cost and fuel model",
    "Monitor — identify levers to push toward Cluster 1",
    "Priority for cost review and maintenance intervention"
  )
)

kable(archetype_df, caption = "Provisional Cluster Archetypes — Focused 5-Feature Result") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = TRUE) %>%
  column_spec(2, bold = TRUE) %>%
  row_spec(3, background = "#fde8e8") %>%
  row_spec(1, background = "#e8f8ee")
```

> **Note:** Provisional names above are based on expected cluster behaviour from the
> EDA findings. Revise them after reading the actual cross-tab plots above, which
> show the true composition of each cluster.

---

# 12. Cluster Stability Check

## Agreement Between All Algorithms

```{r agreement}
# Adjusted Rand Index — measures pairwise agreement between clusterings
ari_km_hc  <- round(adjustedRandIndex(km_result$cluster,          hc_labels), 4)
ari_km_gmm <- round(adjustedRandIndex(km_result$cluster,          gmm_result$classification), 4)
ari_km_pam <- round(adjustedRandIndex(km_result$cluster,          pam_result$clustering), 4)
ari_hc_gmm <- round(adjustedRandIndex(hc_labels,                  gmm_result$classification), 4)
ari_hc_pam <- round(adjustedRandIndex(hc_labels,                  pam_result$clustering), 4)
ari_gmm_pam<- round(adjustedRandIndex(gmm_result$classification,  pam_result$clustering), 4)

ari_df <- data.frame(
  Pair        = c("K-Means vs Hierarchical", "K-Means vs GMM", "K-Means vs PAM",
                  "Hierarchical vs GMM",      "Hierarchical vs PAM", "GMM vs PAM"),
  ARI         = c(ari_km_hc, ari_km_gmm, ari_km_pam,
                  ari_hc_gmm, ari_hc_pam, ari_gmm_pam),
  Agreement   = case_when(
    c(ari_km_hc, ari_km_gmm, ari_km_pam, ari_hc_gmm, ari_hc_pam, ari_gmm_pam) > 0.6 ~ "Strong",
    c(ari_km_hc, ari_km_gmm, ari_km_pam, ari_hc_gmm, ari_hc_pam, ari_gmm_pam) > 0.3 ~ "Moderate",
    TRUE ~ "Weak"
  )
)

kable(ari_df, caption = "Adjusted Rand Index — Pairwise Algorithm Agreement") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  column_spec(2, bold = TRUE) %>%
  column_spec(3, color = ifelse(ari_df$Agreement == "Strong", "darkgreen",
                         ifelse(ari_df$Agreement == "Moderate", "darkorange", "red")),
              bold = TRUE)
```

```{r agreement-plot}
df_agreement <- data.frame(
  KMeans = factor(km_result$cluster),
  HC     = factor(hc_labels),
  GMM    = factor(gmm_result$classification),
  PAM    = factor(pam_result$clustering)
)

p_hc <- ggplot(df_agreement, aes(x = KMeans, fill = HC)) +
  geom_bar(position = "fill", color = "white") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Dark2") +
  labs(title = paste("K-Means vs HC | ARI =", ari_km_hc),
       x = "K-Means", y = "Proportion", fill = "HC") +
  theme_minimal(base_size = 11) + theme(legend.position = "bottom")

p_gmm <- ggplot(df_agreement, aes(x = KMeans, fill = GMM)) +
  geom_bar(position = "fill", color = "white") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = paste("K-Means vs GMM | ARI =", ari_km_gmm),
       x = "K-Means", y = "Proportion", fill = "GMM") +
  theme_minimal(base_size = 11) + theme(legend.position = "bottom")

p_pam <- ggplot(df_agreement, aes(x = KMeans, fill = PAM)) +
  geom_bar(position = "fill", color = "white") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Pastel2") +
  labs(title = paste("K-Means vs PAM | ARI =", ari_km_pam),
       x = "K-Means", y = "Proportion", fill = "PAM") +
  theme_minimal(base_size = 11) + theme(legend.position = "bottom")

grid.arrange(p_hc, p_gmm, p_pam, ncol = 3,
             top = "Pairwise Cluster Agreement vs K-Means")
```

---

# 13. Final Summary

```{r final-summary, echo=FALSE}
best_algo <- c("K-Means", "Hierarchical (Ward.D2)", "DBSCAN",
               "GMM (EM)", "PAM (K-Medoids)")[which.max(all_sil)]

summary_df <- data.frame(
  Step = c(
    "Feature scaling",
    "Categorical encoding",
    "Optimal K selection",
    "Algorithm 1 — K-Means",
    "Algorithm 2 — Hierarchical (Ward.D2)",
    "Algorithm 3 — DBSCAN",
    "Algorithm 4 — GMM (EM)",
    "Algorithm 5 — PAM (K-Medoids)",
    "Best algorithm (by silhouette)",
    "Cluster interpretation"
  ),
  Detail = c(
    "Z-score standardisation (mean=0, SD=1) applied to all 14 features",
    paste("One-hot encoding: Maintenance_Status (", paste(ohe_features[startsWith(ohe_features,"Maintenance")], collapse=", "), ") | Weather (", paste(ohe_features[startsWith(ohe_features,"Weather")], collapse=", "), ")"),
    paste("K =", chosen_k, "selected via Elbow + Silhouette + Gap Statistic"),
    paste("Avg silhouette:", km_avg_sil, "| Hard assignment | Spherical clusters"),
    paste("Avg silhouette:", hc_avg_sil, "| Ward.D2 linkage | Dendrogram cut at K =", chosen_k),
    paste("Avg silhouette: N/A | Clusters found:", n_db_clusters,
          "| Noise:", sum(db_result$cluster == 0)),
    paste("Avg silhouette:", gmm_avg_sil, "| Model:", gmm_result$modelName,
          "| K selected by BIC:", gmm_result$G, "| Soft assignment"),
    paste("Avg silhouette:", pam_avg_sil, "| Hard assignment | Medoid = real voyage | Outlier-robust"),
    paste(best_algo, "— silhouette:", round(max(all_sil, na.rm=TRUE), 4)),
    "Cross-tabs in Section 11 — name each archetype using Ship Type, Route, Engine, Maintenance"
  )
)

kable(summary_df, caption = "Clustering Pipeline — 5-Algorithm Summary") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = TRUE) %>%
  row_spec(which(summary_df$Step == "Best algorithm (by silhouette)"),
           bold = TRUE, background = "#d5f5e3")
```

---

*End of Clustering Report — next step: classifier on `Maintenance_Status`*
