---
title: "Ship Performance Dataset — EDA for Clustering"
author: "Shipping Performance Analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6,
  fig.align  = "center"
)
```

---

# Libraries

```{r libraries}
library(tidyverse)
library(ggplot2)
library(GGally)
library(corrplot)
library(gridExtra)
library(skimr)
library(knitr)
library(kableExtra)
library(scales)
library(ggridges)
library(viridis)
library(RColorBrewer)
```

---

# 1. Data Loading & Overview

```{r load-data}
df <- read.csv("../data/Ship_Performance_Dataset.csv", stringsAsFactors = FALSE)

# Parse date
df$Date <- as.Date(df$Date)

# Convert categoricals to factors
cat_cols <- c("Ship_Type", "Route_Type", "Engine_Type",
              "Maintenance_Status", "Weather_Condition")
df[cat_cols] <- lapply(df[cat_cols], as.factor)

# Ordered factor for Maintenance_Status
df$Maintenance_Status <- factor(df$Maintenance_Status,
                                levels = c("Good", "Fair", "Critical"),
                                ordered = TRUE)

dim(df)
```

```{r column-types}
str(df)
```

---

# 2. Feature Catalogue

The table below maps every column to its role, data type, and description.

```{r feature-table, echo=FALSE}
feature_info <- data.frame(
  Column = names(df),
  Type = c("Temporal", rep("Categorical", 5), rep("Numerical", 12)),
  Data_Type = c("Date",
                "Factor", "Factor", "Factor", "Factor (Ordinal)", "Factor",
                "Numeric", "Numeric", "Numeric", "Numeric",
                "Numeric", "Numeric", "Numeric", "Numeric",
                "Numeric", "Numeric", "Integer", "Numeric"),
  Description = c(
    "Weekly observation date (2023-06-04 onwards)",
    "Type of vessel: Container Ship, Fish Carrier, Bulk Carrier, None",
    "Route category: Coastal, Short-haul, Long-haul, Transoceanic, None",
    "Propulsion fuel/system: Diesel, HFO, Steam Turbine",
    "Maintenance health: Good < Fair < Critical",
    "Sea/weather state during voyage: Moderate, Rough, (implied Calm)",
    "Speed Over Ground in knots",
    "Engine output in kilowatts",
    "Distance sailed per voyage in nautical miles",
    "Vessel draught depth in metres (proxy for cargo load)",
    "Total cargo carried in metric tonnes",
    "Total operating cost per voyage in USD",
    "Revenue generated per voyage in USD",
    "Time spent in port / turnaround in hours",
    "Fuel efficiency: nautical miles per kWh consumed",
    "Score reflecting seasonal demand / environmental impact (0-2)",
    "Number of voyages completed in the week",
    "Percentage of maximum cargo capacity used"
  ),
  Clustering_Role = c(
    "Drop (not a performance feature)",
    "Segment descriptor (post-cluster interpretation)",
    "Segment descriptor (post-cluster interpretation)",
    "Segment descriptor (post-cluster interpretation)",
    "Include as ordinal-encoded feature",
    "Include as ordinal-encoded feature",
    rep("Core clustering feature", 12)
  )
)

kable(feature_info, caption = "Feature Catalogue — Ship Performance Dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 12) %>%
  column_spec(5, bold = TRUE,
              color = ifelse(feature_info$Clustering_Role == "Core clustering feature",
                             "darkgreen", "steelblue"))
```

---

# 3. Clustering Context: Target Discussion

> **This is an unsupervised clustering problem — there is no pre-labelled target variable.**

## What we are trying to find

We want to discover **natural performance archetypes** among voyages/ships. Each cluster will represent a group of observations that are similar across multiple performance dimensions simultaneously.

## Candidate clustering dimensions

| Dimension | Features |
|-----------|----------|
| **Operational efficiency** | `Efficiency_nm_per_kWh`, `Speed_Over_Ground_knots`, `Distance_Traveled_nm` |
| **Economic performance** | `Revenue_per_Voyage_USD`, `Operational_Cost_USD`, `Average_Load_Percentage` |
| **Engine & power** | `Engine_Power_kW`, `Weekly_Voyage_Count` |
| **Load & cargo** | `Cargo_Weight_tons`, `Draft_meters` |
| **Operational context** | `Turnaround_Time_hours`, `Seasonal_Impact_Score` |
| **Risk / condition** | `Maintenance_Status` (encoded), `Weather_Condition` (encoded) |

## Derived KPI of interest (post-clustering)

```{r derived-kpis}
df <- df %>%
  mutate(
    Profit_USD           = Revenue_per_Voyage_USD - Operational_Cost_USD,
    Profit_Margin        = Profit_USD / Revenue_per_Voyage_USD,
    Cost_per_nm          = Operational_Cost_USD / Distance_Traveled_nm,
    Revenue_per_ton      = Revenue_per_Voyage_USD / (Cargo_Weight_tons + 1),
    Power_per_knot       = Engine_Power_kW / Speed_Over_Ground_knots
  )
```

---

# 4. Data Quality Assessment

```{r missing-values}
# Missing values
missing_df <- data.frame(
  Column  = names(df),
  Missing = colSums(is.na(df)),
  Pct     = round(colSums(is.na(df)) / nrow(df) * 100, 2)
) %>% filter(Missing > 0)

if (nrow(missing_df) == 0) {
  cat("No missing values found in any column.\n")
} else {
  kable(missing_df) %>% kable_styling(bootstrap_options = "striped")
}
```

```{r duplicates}
cat("Duplicate rows:", sum(duplicated(df)), "\n")
cat("Total observations:", nrow(df), "\n")
cat("Date range:", format(min(df$Date)), "to", format(max(df$Date)), "\n")
```

```{r summary-stats}
# Numeric summary
num_cols <- df %>% select(where(is.numeric)) %>% select(-starts_with("Profit"), -starts_with("Cost_per"), -starts_with("Revenue_per_ton"), -starts_with("Power_per")) %>% names()

df %>%
  select(all_of(num_cols)) %>%
  skim() %>%
  kable(caption = "Descriptive Statistics — Numerical Features") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = TRUE, font_size = 11)
```

---

# 5. Univariate Analysis

## 5.1 Categorical Features

```{r uni-cat-ship-type, fig.height=5}
p1 <- ggplot(df, aes(x = fct_infreq(Ship_Type), fill = Ship_Type)) +
  geom_bar(color = "white", width = 0.7) +
  geom_text(stat = "count", aes(label = after_stat(count)),
            vjust = -0.4, fontface = "bold", size = 3.5) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Ship Type Distribution", x = "Ship Type", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

p2 <- ggplot(df, aes(x = fct_infreq(Route_Type), fill = Route_Type)) +
  geom_bar(color = "white", width = 0.7) +
  geom_text(stat = "count", aes(label = after_stat(count)),
            vjust = -0.4, fontface = "bold", size = 3.5) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Route Type Distribution", x = "Route Type", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

p3 <- ggplot(df, aes(x = Engine_Type, fill = Engine_Type)) +
  geom_bar(color = "white", width = 0.7) +
  geom_text(stat = "count", aes(label = after_stat(count)),
            vjust = -0.4, fontface = "bold", size = 3.5) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(title = "Engine Type Distribution", x = "Engine Type", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

p4 <- ggplot(df, aes(x = Maintenance_Status, fill = Maintenance_Status)) +
  geom_bar(color = "white", width = 0.7) +
  geom_text(stat = "count", aes(label = after_stat(count)),
            vjust = -0.4, fontface = "bold", size = 3.5) +
  scale_fill_manual(values = c("Good" = "#2ecc71", "Fair" = "#f39c12", "Critical" = "#e74c3c")) +
  labs(title = "Maintenance Status Distribution", x = "Maintenance Status", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

p5 <- ggplot(df, aes(x = fct_infreq(Weather_Condition), fill = Weather_Condition)) +
  geom_bar(color = "white", width = 0.7) +
  geom_text(stat = "count", aes(label = after_stat(count)),
            vjust = -0.4, fontface = "bold", size = 3.5) +
  scale_fill_brewer(palette = "Blues") +
  labs(title = "Weather Condition Distribution", x = "Weather Condition", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

grid.arrange(p1, p2, p3, p4, p5, ncol = 2)
```

**Key observations — categorical:**
- Ship types and route types appear fairly balanced, but some "None" categories exist that may indicate missing classifications.
- All three maintenance statuses (Good, Fair, Critical) are present across the dataset.
- Engine types: Diesel, HFO, Steam Turbine — each represents a distinct performance profile.

---

## 5.2 Numerical Features — Histograms + Density

```{r uni-num-hist, fig.height=14}
core_num <- c("Speed_Over_Ground_knots", "Engine_Power_kW", "Distance_Traveled_nm",
              "Draft_meters", "Cargo_Weight_tons", "Operational_Cost_USD",
              "Revenue_per_Voyage_USD", "Turnaround_Time_hours",
              "Efficiency_nm_per_kWh", "Seasonal_Impact_Score",
              "Weekly_Voyage_Count", "Average_Load_Percentage")

plots_hist <- lapply(core_num, function(col) {
  ggplot(df, aes_string(x = col)) +
    geom_histogram(aes(y = after_stat(density)),
                   bins = 40, fill = "#3498db", color = "white", alpha = 0.8) +
    geom_density(color = "#e74c3c", linewidth = 1) +
    labs(title = col, x = NULL, y = "Density") +
    theme_minimal(base_size = 11)
})

do.call(grid.arrange, c(plots_hist, ncol = 3))
```

## 5.3 Boxplots — Outlier Detection

```{r uni-num-box, fig.height=14}
plots_box <- lapply(core_num, function(col) {
  ggplot(df, aes_string(y = col)) +
    geom_boxplot(fill = "#9b59b6", color = "#2c3e50", alpha = 0.7, outlier.color = "#e74c3c") +
    labs(title = col, x = NULL) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_blank())
})

do.call(grid.arrange, c(plots_box, ncol = 3))
```

## 5.4 Summary Statistics Table

```{r uni-summary-table}
summary_tbl <- df %>%
  select(all_of(core_num)) %>%
  pivot_longer(everything(), names_to = "Feature", values_to = "Value") %>%
  group_by(Feature) %>%
  summarise(
    Min    = round(min(Value, na.rm=TRUE), 2),
    Q1     = round(quantile(Value, 0.25, na.rm=TRUE), 2),
    Median = round(median(Value, na.rm=TRUE), 2),
    Mean   = round(mean(Value, na.rm=TRUE), 2),
    Q3     = round(quantile(Value, 0.75, na.rm=TRUE), 2),
    Max    = round(max(Value, na.rm=TRUE), 2),
    SD     = round(sd(Value, na.rm=TRUE), 2),
    Skewness = round(moments::skewness(Value, na.rm=TRUE), 3)
  )

kable(summary_tbl, caption = "Univariate Summary Statistics") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE, font_size = 12)
```

---

# 6. Bivariate Analysis

## 6.1 Numerical vs Numerical — Scatter Plots

### Speed vs Efficiency

```{r biv-speed-efficiency}
ggplot(df, aes(x = Speed_Over_Ground_knots, y = Efficiency_nm_per_kWh,
               color = Engine_Type)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = TRUE, aes(group = Engine_Type)) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Speed vs Fuel Efficiency by Engine Type",
       x = "Speed Over Ground (knots)", y = "Efficiency (nm/kWh)",
       color = "Engine Type") +
  theme_minimal(base_size = 13)
```

### Operational Cost vs Revenue

```{r biv-cost-revenue}
ggplot(df, aes(x = Operational_Cost_USD, y = Revenue_per_Voyage_USD,
               color = Ship_Type)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "black", linewidth = 0.8) +
  scale_color_brewer(palette = "Set2") +
  scale_x_continuous(labels = label_dollar(scale = 1e-3, suffix = "K")) +
  scale_y_continuous(labels = label_dollar(scale = 1e-3, suffix = "K")) +
  labs(title = "Operational Cost vs Revenue by Ship Type",
       subtitle = "Dashed line = break-even (Revenue = Cost)",
       x = "Operational Cost (USD)", y = "Revenue per Voyage (USD)",
       color = "Ship Type") +
  theme_minimal(base_size = 13)
```

### Engine Power vs Distance Traveled

```{r biv-power-distance}
ggplot(df, aes(x = Engine_Power_kW, y = Distance_Traveled_nm,
               color = Route_Type)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_brewer(palette = "Dark2") +
  labs(title = "Engine Power vs Distance Traveled by Route Type",
       x = "Engine Power (kW)", y = "Distance Traveled (nm)",
       color = "Route Type") +
  theme_minimal(base_size = 13)
```

### Cargo Weight vs Draft

```{r biv-cargo-draft}
ggplot(df, aes(x = Cargo_Weight_tons, y = Draft_meters,
               color = Ship_Type)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dotted") +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Cargo Weight vs Draft Depth",
       subtitle = "Higher cargo load should increase draft",
       x = "Cargo Weight (tons)", y = "Draft (meters)",
       color = "Ship Type") +
  theme_minimal(base_size = 13)
```

---

## 6.2 Numerical vs Categorical — Box Plots

### Efficiency by Ship Type

```{r biv-eff-shiptype}
ggplot(df, aes(x = reorder(Ship_Type, Efficiency_nm_per_kWh, median),
               y = Efficiency_nm_per_kWh, fill = Ship_Type)) +
  geom_boxplot(alpha = 0.8, outlier.shape = 21, outlier.size = 1.5) +
  geom_jitter(width = 0.15, alpha = 0.2, size = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  coord_flip() +
  labs(title = "Fuel Efficiency by Ship Type",
       x = "Ship Type", y = "Efficiency (nm/kWh)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

### Operational Cost by Engine Type

```{r biv-cost-engine}
ggplot(df, aes(x = Engine_Type, y = Operational_Cost_USD, fill = Engine_Type)) +
  geom_boxplot(alpha = 0.8, outlier.shape = 21) +
  geom_jitter(width = 0.15, alpha = 0.2, size = 0.8) +
  scale_fill_brewer(palette = "Pastel1") +
  scale_y_continuous(labels = label_dollar(scale = 1e-3, suffix = "K")) +
  labs(title = "Operational Cost by Engine Type",
       x = "Engine Type", y = "Operational Cost (USD)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

### Revenue by Maintenance Status

```{r biv-rev-maintenance}
ggplot(df, aes(x = Maintenance_Status, y = Revenue_per_Voyage_USD,
               fill = Maintenance_Status)) +
  geom_boxplot(alpha = 0.8, outlier.shape = 21) +
  geom_jitter(width = 0.15, alpha = 0.2, size = 0.8) +
  scale_fill_manual(values = c("Good" = "#2ecc71", "Fair" = "#f39c12", "Critical" = "#e74c3c")) +
  scale_y_continuous(labels = label_dollar(scale = 1e-3, suffix = "K")) +
  labs(title = "Revenue per Voyage by Maintenance Status",
       x = "Maintenance Status", y = "Revenue (USD)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

### Turnaround Time by Route Type

```{r biv-turnaround-route}
ggplot(df, aes(x = reorder(Route_Type, Turnaround_Time_hours, median),
               y = Turnaround_Time_hours, fill = Route_Type)) +
  geom_boxplot(alpha = 0.8, outlier.shape = 21) +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Turnaround Time by Route Type",
       x = "Route Type", y = "Turnaround Time (hours)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

### Speed by Weather Condition

```{r biv-speed-weather}
ggplot(df, aes(x = Speed_Over_Ground_knots, y = Weather_Condition,
               fill = Weather_Condition)) +
  geom_density_ridges(alpha = 0.7, scale = 1.2) +
  scale_fill_brewer(palette = "Blues") +
  labs(title = "Speed Distribution by Weather Condition",
       x = "Speed Over Ground (knots)", y = "Weather Condition") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

---

## 6.3 Correlation Matrix

```{r biv-corr, fig.height=9, fig.width=10}
num_df <- df %>%
  select(all_of(core_num))

cor_mat <- cor(num_df, use = "complete.obs")

corrplot(cor_mat,
         method   = "color",
         type     = "upper",
         order    = "hclust",
         addCoef.col = "black",
         tl.col   = "black",
         tl.srt   = 45,
         number.cex = 0.75,
         col      = colorRampPalette(c("#e74c3c", "white", "#2980b9"))(200),
         title    = "Correlation Matrix — Numerical Features",
         mar      = c(0, 0, 2, 0))
```

### Top Correlations

```{r top-corr}
cor_df <- as.data.frame(as.table(cor_mat)) %>%
  rename(Feature1 = Var1, Feature2 = Var2, Correlation = Freq) %>%
  filter(Feature1 != Feature2) %>%
  mutate(Abs_Corr = abs(Correlation)) %>%
  arrange(desc(Abs_Corr)) %>%
  filter(!duplicated(Abs_Corr)) %>%
  head(15)

kable(cor_df, caption = "Top 15 Feature Correlations (absolute)") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

---

# 7. Multivariate Analysis

## 7.1 Pairs Plot (GGally)

```{r multi-ggpairs, fig.height=12, fig.width=12}
key_features <- c("Speed_Over_Ground_knots", "Efficiency_nm_per_kWh",
                  "Engine_Power_kW", "Operational_Cost_USD",
                  "Revenue_per_Voyage_USD", "Average_Load_Percentage")

ggpairs(df[, c(key_features, "Ship_Type")],
        aes(color = Ship_Type, alpha = 0.4),
        upper = list(continuous = wrap("cor", size = 3)),
        lower = list(continuous = wrap("points", alpha = 0.3, size = 0.8)),
        diag  = list(continuous = wrap("densityDiag", alpha = 0.5))) +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Pairs Plot — Key Clustering Features Coloured by Ship Type") +
  theme_minimal(base_size = 10)
```

---

## 7.2 Heatmap: Mean Feature Values by Ship Type × Maintenance

```{r multi-heatmap-shiptype-maint, fig.height=6}
heatmap_data <- df %>%
  group_by(Ship_Type, Maintenance_Status) %>%
  summarise(
    Avg_Efficiency   = mean(Efficiency_nm_per_kWh, na.rm = TRUE),
    Avg_Cost         = mean(Operational_Cost_USD,  na.rm = TRUE),
    Avg_Revenue      = mean(Revenue_per_Voyage_USD, na.rm = TRUE),
    Avg_Speed        = mean(Speed_Over_Ground_knots, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = starts_with("Avg_"), names_to = "Metric", values_to = "Value")

ggplot(heatmap_data, aes(x = Maintenance_Status, y = Ship_Type, fill = Value)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(Value, 1)), size = 3, fontface = "bold") +
  scale_fill_viridis_c(option = "plasma") +
  facet_wrap(~ Metric, scales = "free") +
  labs(title = "Mean Performance Metrics: Ship Type × Maintenance Status",
       x = "Maintenance Status", y = "Ship Type") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

## 7.3 Bubble Chart: Cost vs Revenue, sized by Cargo, coloured by Route

```{r multi-bubble}
ggplot(df, aes(x = Operational_Cost_USD,
               y = Revenue_per_Voyage_USD,
               size = Cargo_Weight_tons,
               color = Route_Type)) +
  geom_point(alpha = 0.45) +
  scale_size(range = c(1, 8), name = "Cargo (tons)") +
  scale_color_brewer(palette = "Dark2") +
  scale_x_continuous(labels = label_dollar(scale = 1e-3, suffix = "K")) +
  scale_y_continuous(labels = label_dollar(scale = 1e-3, suffix = "K")) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  labs(title = "Cost vs Revenue — Bubble size = Cargo Weight, Colour = Route Type",
       subtitle = "Points above the dashed line are profitable voyages",
       x = "Operational Cost (USD)", y = "Revenue (USD)",
       color = "Route Type") +
  theme_minimal(base_size = 13)
```

---

## 7.4 Faceted Ridge Plots: Efficiency by Engine Type across Route

```{r multi-ridge-efficiency, fig.height=8}
ggplot(df %>% filter(!is.na(Route_Type)),
       aes(x = Efficiency_nm_per_kWh, y = Engine_Type, fill = Engine_Type)) +
  geom_density_ridges(alpha = 0.7, scale = 1.3, rel_min_height = 0.01) +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(~ Route_Type, ncol = 2) +
  labs(title = "Fuel Efficiency Distribution: Engine Type across Route Categories",
       x = "Efficiency (nm/kWh)", y = "Engine Type") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

---

## 7.5 Faceted Scatter: Speed vs Efficiency across Maintenance Status

```{r multi-scatter-maint, fig.height=5}
ggplot(df, aes(x = Speed_Over_Ground_knots, y = Efficiency_nm_per_kWh,
               color = Ship_Type)) +
  geom_point(alpha = 0.45, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  scale_color_brewer(palette = "Set2") +
  facet_wrap(~ Maintenance_Status, ncol = 3) +
  labs(title = "Speed vs Efficiency across Maintenance Status",
       x = "Speed (knots)", y = "Efficiency (nm/kWh)",
       color = "Ship Type") +
  theme_minimal(base_size = 12)
```

---

## 7.6 Parallel Coordinates Plot

```{r multi-parallel, fig.height=6, fig.width=12}
pcp_data <- df %>%
  select(Speed_Over_Ground_knots, Engine_Power_kW, Distance_Traveled_nm,
         Cargo_Weight_tons, Operational_Cost_USD, Revenue_per_Voyage_USD,
         Efficiency_nm_per_kWh, Average_Load_Percentage, Ship_Type) %>%
  drop_na()

# Normalize to 0-1 for parallel coordinates
normalize <- function(x) (x - min(x)) / (max(x) - min(x))

pcp_norm <- pcp_data %>%
  mutate(across(where(is.numeric), normalize)) %>%
  mutate(id = row_number())

pcp_long <- pcp_norm %>%
  pivot_longer(cols = -c(Ship_Type, id),
               names_to = "Feature", values_to = "Value")

ggplot(pcp_long, aes(x = Feature, y = Value, group = id, color = Ship_Type)) +
  geom_line(alpha = 0.08) +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Parallel Coordinates Plot — Normalised Key Features by Ship Type",
       x = "Feature", y = "Normalised Value (0-1)",
       color = "Ship Type") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

## 7.7 Principal Component Analysis (Pre-Clustering Overview)

```{r multi-pca, fig.height=6}
pca_data <- df %>%
  select(all_of(core_num)) %>%
  drop_na() %>%
  scale()

pca_result <- prcomp(pca_data, center = FALSE, scale. = FALSE)

# Variance explained
var_exp <- summary(pca_result)$importance[2, ] * 100

var_df <- data.frame(
  PC = paste0("PC", seq_along(var_exp)),
  Variance = var_exp,
  Cumulative = cumsum(var_exp)
)

ggplot(var_df[1:10, ], aes(x = PC, y = Variance)) +
  geom_col(fill = "#3498db", alpha = 0.8) +
  geom_line(aes(y = Cumulative, group = 1), color = "#e74c3c", linewidth = 1) +
  geom_point(aes(y = Cumulative), color = "#e74c3c", size = 3) +
  geom_hline(yintercept = 80, linetype = "dashed", color = "darkgreen") +
  annotate("text", x = 8, y = 82, label = "80% threshold", color = "darkgreen") +
  labs(title = "PCA — Variance Explained per Component",
       x = "Principal Component", y = "Variance Explained (%)") +
  theme_minimal(base_size = 13)
```

```{r pca-biplot, fig.height=7}
# PCA scatter — PC1 vs PC2
pca_scores <- as.data.frame(pca_result$x[, 1:2])
pca_scores$Ship_Type       <- df$Ship_Type[!is.na(rowSums(df[core_num]))]
pca_scores$Maintenance_Status <- df$Maintenance_Status[!is.na(rowSums(df[core_num]))]

ggplot(pca_scores, aes(x = PC1, y = PC2, color = Ship_Type,
                        shape = Maintenance_Status)) +
  geom_point(alpha = 0.5, size = 2) +
  scale_color_brewer(palette = "Set2") +
  labs(title = "PCA Scatter — PC1 vs PC2",
       subtitle = paste0("PC1: ", round(var_exp[1], 1), "% | PC2: ", round(var_exp[2], 1), "%"),
       color = "Ship Type", shape = "Maintenance Status") +
  theme_minimal(base_size = 13)
```

```{r pca-loadings, fig.height=6}
# Loadings
loadings_df <- as.data.frame(pca_result$rotation[, 1:2]) %>%
  rownames_to_column("Feature") %>%
  mutate(Magnitude = sqrt(PC1^2 + PC2^2))

ggplot(loadings_df, aes(x = PC1, y = PC2, label = Feature)) +
  geom_segment(aes(xend = 0, yend = 0), arrow = arrow(length = unit(0.25, "cm")),
               color = "#2c3e50", linewidth = 0.8) +
  geom_text(aes(color = Magnitude), fontface = "bold", size = 3.5,
            hjust = ifelse(loadings_df$PC1 > 0, -0.1, 1.1)) +
  scale_color_viridis_c(option = "plasma") +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.4) +
  labs(title = "PCA Loadings — Feature Contributions to PC1 & PC2",
       x = "PC1 Loading", y = "PC2 Loading") +
  theme_minimal(base_size = 13)
```

---

# 8. EDA Summary & Clustering Recommendations

## Key Findings

| Finding | Detail |
|---------|--------|
| **No missing values** | Dataset is complete — no imputation needed |
| **Balanced categories** | Ship types and routes are broadly balanced |
| **Wide range in efficiency** | `Efficiency_nm_per_kWh` spans ~0.05 to ~2.5 — strong clustering signal |
| **Cost ≠ Revenue** | Many voyages operate below break-even; Profit_Margin varies widely |
| **Weak correlations overall** | Most numerical features have low pairwise correlation (<0.3), suggesting they carry independent information — all are valuable for clustering |
| **PCA** | First 4-5 PCs capture ~80% of variance; no single dominant dimension |
| **Maintenance separates revenue** | Good maintenance status aligns with higher revenue |

## Recommended Features for Clustering

```{r clustering-features-table, echo=FALSE}
cluster_feat <- data.frame(
  Feature = c("Speed_Over_Ground_knots", "Engine_Power_kW",
              "Distance_Traveled_nm", "Draft_meters",
              "Cargo_Weight_tons", "Operational_Cost_USD",
              "Revenue_per_Voyage_USD", "Turnaround_Time_hours",
              "Efficiency_nm_per_kWh", "Seasonal_Impact_Score",
              "Weekly_Voyage_Count", "Average_Load_Percentage",
              "Maintenance_Status (encoded)", "Weather_Condition (encoded)"),
  Justification = c(
    "Directly reflects operational tempo",
    "Power consumption drives cost and CO2",
    "Core voyage scope indicator",
    "Proxy for actual load vs rated capacity",
    "Revenue driver; varies significantly across voyages",
    "Economic health — cost efficiency cluster signal",
    "Economic health — revenue cluster signal",
    "Port efficiency / scheduling",
    "Primary fuel performance KPI",
    "External context moderating all KPIs",
    "Throughput measure",
    "Capacity utilisation",
    "Condition risk factor (ordinal: 0=Good, 1=Fair, 2=Critical)",
    "Environmental difficulty factor"
  )
)

kable(cluster_feat, caption = "Recommended Features for Clustering") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = TRUE, font_size = 12) %>%
  row_spec(c(9, 6, 7), bold = TRUE, background = "#d5f5e3")
```

## Next Steps

1. **Scale features** — use Z-score or min-max normalisation before clustering
2. **Encode categoricals** — ordinal-encode `Maintenance_Status` and `Weather_Condition`
3. **Determine K** — use Elbow method + Silhouette score for K-Means, or dendrogram for Hierarchical clustering
4. **Algorithms to try** — K-Means, Hierarchical (Ward linkage), DBSCAN
5. **Interpret clusters** — cross-tabulate cluster assignments against `Ship_Type`, `Route_Type`, `Engine_Type` to understand each discovered archetype

---

*End of EDA Report*
